---
title: "Assignment 5 - Data Visualization (Incarceration)"
author: "Ryder Forsythe"
date: "2024-05-27"
output: html_document
---

## Introduction

I chose to look at the percentage of incarcerated individuals who are not white, and compared that to the percentage of nonwhite individuals between ages 15 and 64 in the general population. The total percentage of incarcerated individuals who are not white in 2016, the most recent year with complete data, is 60.25%, far higher than the percentage of nonwhite people between the ages of 15 and 64 in the general population, which in 2016 was 39.67%. This percentage is highest in the state of California, at 78.48%. This percentage has decreased by 17.5% since 1990, when 77.6% of the incarcerated population was not white, a staggeringly higher percentage than in the general population, which was only 24.25% nonwhite. This indicates a dramatic, systemic bias in which people were incarcerated - one that has only somewhat abated since then.

## Data

These datasets were collated and published by the Vera Institute of Justice, using public datasets from the United States Department of Justice Bureau of Justice Statistics (BJS), as well as from various state departments of correction where federal data was not available or incomplete. These agencies collected data through a variety of surveys and sources, including the National Corrections Reporting Program, the Annual Survey of Jails, the Census of Jails, the Uniform Crime Reporting Program, National Vital Statistics System Data from the Department of Agriculture, and the Death in Custody Reporting Program. All together, this dataset presents as complete a look as possible at race and ethnicity in prisons and jails in the United States from 1970 to 2018, aggregated by county. In the Prison Population (prison_pop) dataset, which I focused on, there are 153811 rows and 37 columns. This dataset intersects two very important societal areas of interest - mass incarceration and systemic racial discrimination. It is therefore very important that this be as accurate to the real world as possible, because it demonstrates a clear set of systemic biases towards the incarceration of nonwhite people. Several states do not have counties with recorded prison populations; racial data within prisons is only available from 1990 on, and it is unclear whether any counties with prisons have been excluded due to a lack of data.

## Trends Over Time

```{r}
# load packages and data
library(tidyverse)
library(scales)

prison_pop <- read.csv(url("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv"))

# create new variables and set up new dataframe grouped by year
prison_pop_by_year <- prison_pop %>% 
  group_by(year) %>%
  summarize(
    prison_percent_nonwhite = 1 - sum(white_prison_pop, na.rm = TRUE) / sum(total_prison_pop, na.rm = TRUE),
    total_percent_nonwhite = 1 - sum(white_pop_15to64, na.rm = TRUE)/sum(total_pop_15to64, na.rm = TRUE)
  )

# create line graph
legend_colors <- c("Prison Nonwhite %" = "coral3", "Total Nonwhite %" = "dodgerblue3")
chart1 <- filter(prison_pop_by_year, year > 1989 & year < 2017) %>% ggplot(aes(x = year)) +
  geom_line(aes(y = prison_percent_nonwhite, color = "Prison Nonwhite %"), linewidth = 1) +
  geom_line(aes(y = total_percent_nonwhite, color = "Total Nonwhite %"), linewidth = 1) +
  coord_cartesian(xlim = c(1990, 2016), ylim = c(0, 1)) +
  labs(
    x = "Year",
    title = "Nationwide Nonwhite Percentage of Prison and Total Population, 1990-2016",
    color = "Legend"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(from = 1990, to = 2016, by = 5)) +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = legend_colors)

chart1
```

I wanted to look at how the nonwhite proportion of incarcerated individuals has changed over time as the racial makeup of the general population has changed. Counterintuitively, while the United States has become less white over time, the proportion of incarcerated individuals who are white has increased. This could be explained by a systemic bias towards the incarceration of nonwhite people decreasing over time; as the chart indicates, nonwhite people are still very disproportionately incarcerated - by a factor of more than 2 to 1 in 2016 - but the bias seems to be becoming less pronounced.

## Variable Comparison

```{r}

# isolate 2016
prison_pop_2016 <- prison_pop %>% filter(year == "2016") %>%
  mutate(nonwhite_total_percentage = 1 - white_pop_15to64/total_pop_15to64,
         nonwhite_prison_percentage = 1 - white_prison_pop/total_prison_pop
         )
# create scatterplot
prison_pop_2016 %>% ggplot(aes(x = nonwhite_total_percentage, y = nonwhite_prison_percentage)) +
  geom_point(aes(size = total_prison_pop), alpha = 0.3) +
  labs(
    x = "Nonwhite % of County Population",
    y = "Nonwhite % of Prison Population",
    title = "Nonwhite % of Total Population vs Prison Population, by County, 2016",
    size = "# of Incarcerated Individuals"
  ) +
  theme_minimal() +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent)
```

This plot shows the the nonwhite percentage of a county's population (from ages 15 to 64) plotted against the percentage of incarcerated people in that county who are nonwhite, as of 2016. What this shows is that there is a clear correlation between the nonwhite general population and the nonwhite prison population, but also that this relationship isn't quite linear. As the nonwhite percentage of a county's population increases, the nonwhite percentage of their incarcerated population generally increases much faster, to the point where a very large number of majority-white counties have majority-nonwhite prisons. This relationship becomes less extreme as the county becomes more nonwhite - it looks more like a logarithmic relationship than a linear one. The points on the scatterplot are also sized based on the absolute number of incarcerated people in each county - this is to illustrate the fact that although the vast majority of counties are very white, the largest counties are generally much less white and have many orders of magnitude more people.

## Map

```{r}
# load packages and data
library(tidyverse)
library(scales)
library(sf)
library(maps)

prison_pop <- read.csv(url("https://raw.githubusercontent.com/melaniewalsh/Neat-Datasets/main/us-prison-pop.csv"))

states <- map_data("state")

# reformat states to match map_data package defaults
prison_pop$state <- prison_pop$state %>% 
  case_match("AL" ~ "alabama",
             "AK" ~ "alaska",
             "AZ" ~ "arizona", 
             "AR" ~ "arkansas",
             "CA" ~ "california",
             "CO" ~ "colorado",
             "CT" ~ "connecticut",
             "DE" ~ "delaware",
             "DC" ~ "district of columbia",
             "FL" ~ "florida",
             "GA" ~ "georgia",
             "HI" ~ "hawaii",
             "ID" ~ "idaho",
             "IL" ~ "illinois", 
             "IN" ~ "indiana",
             "IA" ~ "iowa",
             "KS" ~ "kansas",
             "KY" ~ "kentucky",
             "LA" ~ "louisiana",
             "ME" ~ "maine",
             "MD" ~ "maryland",
             "MA" ~ "massachusetts",
             "MI" ~ "michigan",
             "MN" ~ "minnesota",
             "MS" ~ "mississippi",
             "MO" ~ "missouri", 
             "MT" ~ "montana",
             "NE" ~ "nebraska",
             "NV" ~ "nevada",
             "NH" ~ "new hampshire",
             "NJ" ~ "new jersey",
             "NM" ~ "new mexico",
             "NY" ~ "new york",
             "NC" ~ "north carolina",
             "ND" ~ "north dakota",
             "OH" ~ "ohio",
             "OK" ~ "oklahoma", 
             "OR" ~ "oregon",
             "PA" ~ "pennsylvania",
             "RI" ~ "rhode island",
             "SC" ~ "south carolina",
             "SD" ~ "south dakota",
             "TN" ~ "tennessee",
             "TX" ~ "texas",
             "UT" ~ "utah",
             "VT" ~ "vermont",
             "VA" ~ "virginia",
             "WA" ~ "washington",
             "WV" ~ "west virginia",
             "WI" ~ "wisconsin",
             "WY" ~ "wyoming"
               ) 
# set up new dataframe for merging
prison_pop_mapping <- prison_pop %>% 
  filter(year == 2016) %>%
  mutate(region = state) %>%
  group_by(region) %>%
  summarize(nonwhite_prison_percentage = 1 - sum(white_prison_pop, na.rm = TRUE)/sum(total_prison_pop, na.rm = TRUE), 
            nonwhite_total_percentage = 1 - sum(white_pop_15to64, na.rm = TRUE)/sum(total_pop_15to64, na.rm = TRUE))

# merge state shapes with prison dataset
map_df <- merge(prison_pop_mapping, states, by = "region")
  
# create the map
map_df %>% ggplot(aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = nonwhite_prison_percentage)) + 
  coord_map("albers", lat0 = 45.5, lat1 = 29.5) +
  theme_minimal() +
  labs(
    x = "longitude",
    y = "latitude",
    title = "Nonwhite Percentage of Incarcerated Population by State, 2016",
    fill = "Nonwhite % of Prison Population"
  ) +
  scale_fill_viridis_c()
```

This map depicts the nonwhite percentage of incarcerated people by state in 2016. There are some clear regional trends in the dataset: New York and New Jersey, both situated in the New York City metropolitan area, have very nonwhite prison populations, as does California; smaller and more rural states generally have a whiter prison population - see Maine, Vermont, Kentucky, and Wyoming. However, the South appears to have a much higher proportion of nonwhite people behind bars than its neighbors, despite these states being generally as whiter and more rural than, for example, Nevada. This indicates some degree of regional policy, culture, or socioeconomic factors that might influence the way that nonwhite people are treated by the judicial system.
